---
title: "Case Analysis 1"
subtitle: "Home Rule and the ROT"
date: "2023-11-12"
---

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
library(dplyr)
library(tibble)
library(pander)
```

## The memo
Please prepare a memo responding to the "case" found below. Follow the standard format from PSPA 600. If you have any questions about the format, please consult the Memo Resources folder in the Assignment Resources folder on the main Blackboard page. If you have consulted these resources and still have further questions, please email me. 

Be sure to thoroughly read the "case" and prepare a memo that satisfies all the requirements. You are limited to 1000 words in the main document (about two pages, single-spaced) plus a (hopefully less than 1 page) executive summary. Please include any citations as footnotes (the exact citation style is up to you (the department requires Chicago/Turabian) as long as it is consistent throughout). These will not count against your word limit.

## The case
On November 8, 2022, the [Village of Barrington, IL](https://www.barrington-il.gov/), adopted home rule status via referendum. Barrington, split between Cook and Lake counties, voted approximately 54% in favor of home rule status. 

The village president and trustees have indicated that they will exercise their new home rule powers, not to increase property taxes, but to adopt a home rule sales tax of one percent (Barrington does not currently levy a non-home rule sales tax). The stated goal of this tax plan is to raise additional funds for infrastructure improvements and have non-residents pay some of those funds. 

Village Manager Scott Anderson is in favor of this plan; however, he is concerned about all the potential equity and neutrality issues at play. He has asked you to prepare a memo briefing him on the equity and neutrality issues associated with adopting a new sales tax in Barrington. Specifically, what are the equity and neutrality issues, what are the tradeoffs between them, how do these issues play out in Barrington specifically, and, given this information, should the trustees reconsider their plan?

### Context (or things to think about while answering the case)
Memo 1 deals with the interplay of home rule and sales taxation. There are many ways to answer this question, but I encourage you to give equal time to equity and neutrality. Consider who lives in Barrington (this Census website can help, [Barrington IL Quick Facts](https://www.census.gov/quickfacts/fact/table/barringtonvillageillinois,US/PST045222)). Some equity concerns may be moot if specific segments of the income distribution are absent. Similarly, think about the businesses that Barrington calls home. Which ones might be most heavily impacted by a change in sales taxes, and in what ways might this influence neutrality?

Furthermore, I am asking you to refrain from answering the memo questions theoretically. Elected officials and managers rarely want the theoretical answer. They want to know what concerns exist and how those might impact outcomes, so take a little time to familiarize yourself with Barrington’s local context. Examine the facts on the ground and provide commentary to Manager Anderson. 

### Background Information

```{r sales-tax, include=FALSE}
revenue = tribble(
    ~Category , ~`Cook County`, ~`Lake County`,
    "General Merchandise", 1207, 242,
    "Food", 42511, 589833,
    "Drinking & Eating Places", 173970, 137030,
    "Apparel", 33634, 23638,
    "Furniture & Household & Radio", 48406, 30638,
    "Lumber, Building, Hardware", 102653, 5572,
    "Automotive & Filling Stations", 4105594, 1491312,
    "Drugs & Misc. Retail", 335802, 413918,
    "Agriculture & All Other", 128781, 194005,
    "Manufacturers", 20781, 8288
) |>
  mutate(
    Total = `Cook County` + `Lake County`,
    #Total = scales::label_dollar(accuracy = 1)(Total)
  )

```

```{r}
#| label: tbl-revenues
#| echo: false
#| results: asis
#| warning: false
#| tbl-cap: Revenue Generated by 1% Municipal Tax in Barrington, IL (FY2022)

revenue_total = tibble(Category = "Total",
                            `Cook County` = sum(revenue$`Cook County`),
                            `Lake County` = sum(revenue$`Lake County`),
                            Total = sum(revenue$Total))

bind_rows(revenue, revenue_total) |> 
  mutate(across(`Cook County`:Total, ~ scales::label_dollar(accuracy = 1)(.))) |>
  select(Category, `Cook County`, `Lake County`, Total) |>
  pandoc.table(justify = "lccc", missing = "—", digits = c(1, 3))


```

Source: <https://www.revenue.state.il.us/app/kob/index.jsp>


Note: You can look up other (neighboring) municipalities at the link above. The comparable column is (MT).

You can compare current tax rates using the IL Tax Finder website (note: Barrington and other nearby municipalities might be in multiple counties).